The effects of context inference on motor learning: savings, de-adaptation and spontaneous recovery

### 2021-09-30 adaptation/article/article.tex
Update on Overleaf.

\setlength{\parskip}{1em}
\def \fref #1{Figure \ref{#1}}     % Reference figures
\def \tref #1{Table \ref{#1}}      % Reference tables
\def \eref #1{Equation \ref{#1}}   % Reference equations
\def \sref #1{Section '\nameref{#1}'}    % Reference sections
\def \supmat {the Supplementary Materials}

@@

\abstract{Abstract}
Humans have been shown to adapt their movements when a sudden change to the dynamics of the environment is introduced, a phenomenon called motor adaptation. If the change is reverted, the adaptation is also quickly reverted. Human are also able to adapt to multiple changes in dynamics presented separately, and to be able to switch between adapted movements on the fly. Such switching relies on contextual information which is often noisy or misleading, which affects the switch between adaptations. In this work, we focus on the behavioral phenomena effected by imperfect contextual information and present a hierarchical computational model for motor adaptation based on exact Bayesian inference, which explicitly takes into account contextual information and how the dynamics of context inference affect learning and action selection. We show how our model provides a unifying explanation for the effects of sensory cues and proprioceptive information on switching between tasks, the effects of environmental volatility and predictability on learning, and the effects of previously-learned adaptations on switching between tasks, in addition to the well studied savings, de-adaptation and spontaneous recovery.

\section{Introduction}
Humans are able to adapt movements to changing dynamics of the environment and of the body itself. We adapt to slow changes in our body dynamics, e.g. because of muscle exhaustion throughout the day, but we can also adapt to sudden changes in the environment. For example, when picking up a heavy object, we must adapt our motor commands to be able to move the extra weight safely. Even before picking it up, the size and material of an object provide us with contextual information which we can use to prepare the correct posture for moving the object. However, contextual information cannot always be relied on, e.g. a big object of unknown material.

It has been shown that humans can learn multiple tasks during the course of a single experiment, as well as learn to adapt to often-opposing modifications to the dynamics of movements \citep[e.g.][]{Gandolfo_Motor_1996,Shadmehr_Functional_1997}, although interference between tasks can slow down learning \citep[e.g.][]{Brashers-Krug_Consolidation_1996,Sing_Reduction_2010}. Additionally, humans have been shown to dynamically switch between different learned adaptation \citep[e.g.][]{Davidson_Scaling_2004,Ethier_Spontaneous_2008,Lee_Dual_2009}. The mechanisms behind these switches, the specifics of how these switches manifest in behavior, and the ways in which switching can be impeded or improved, are still an active topic of research.

@@

By introducing blocks of trials in which a manipulation is present (e.g. a mechanical arm exerts a force), experimenters are able to observe the dynamics of adaptation through the lens of motor error. A number of phenomena have been observed that persist across participants and across many different motor adaptation experiments: savings, quick de-adaptation, anterograde interference and spontaneous recovery.

To explain these phenomena, a number of computational models have been introduced. The least complex of these models are linear learners \citep[e.g.][]{Smith_Interacting_2006,Forano_Timescales_2020,Scheidt_Learning_2001}, in which a single one-dimensional variable is used to describe the adaptation of motor commands given the observed motor errors. By having multiple time scales, each with a learning rate and a forgetting rate, these models have been fitted to experimental data, partially explaining these distinct experimental phenomena.

Bayesian accounts of motor adaptation have also been presented, providing an alternative explanation for savings and quick de-adaptation in the form of switching between internal models of adaptation via predictions generated by these internal models \citep{Kording_Bayesian_2004,Oh_Minimizing_2019}. These models replace the idea of explicit multiple time scales with an estimation of their own uncertainty: as the model practices the same task, the uncertainty on its estimate of the underlying modification is reduced, which lowers variability in responses and increases the stability of the estimate.

@@

In this work, we show that these unexplained phenomena, alongside savings, de-adaptation, anterograde interference and spontaneous recovery, can be explained in terms of the concept of context, an idea that has recently been gaining traction in different fields \cite[e.g.][]{Sanders_Hippocampal_2020,Hunter_Contextsensitive_2021}. In mechanical arm experiments, we can think of one context in which the arm exerts no force and one in which it does. By inferring the current context, a participant can know how to counteract the mechanical arm and succeed in a reaching movement. In some experiments, the context is clearly inferable via sensory cues \cite[e.g.][]{Gandolfo_Motor_1996,Shadmehr_Adaptive_1994,Ethier_Spontaneous_2008}. In many other experiments, the context is not directly observable, and context inference takes the form of an evidence-accumulating process. In this work, we propose that this process can be observed in behavior and can be used to explain the phenomena outlined above.

To this end, we present a computational model for motor adaptation inspired by models such as MOSAIC \citep{Wolpert_Multiple_1998} and Bayesian switching models \citep{Kording_Bayesian_2004,Oh_Minimizing_2019}. We expand on these models by expanding the concept of context and introducing an explicit mechanism for inferring the context based on Bayesian inference. In our model, the context encompasses all the information relevant to making decisions, both environmental and internal to the agent making these decisions. To infer the current context, our model integrates several sources of information, including the predictions made by its internal forward models, an estimate of the reliability of these predictions, direct sensory information (such as visual contextual cues), proprioceptive information, and feedback provided by the environment. Having inferred the context, the model can choose the best-fitting internal model with which to make decisions.

We show that the dynamics of context inference provide a unifying explanation for all the experimental phenomena outlined above, providing a single account for motor adaptation under changing contexts, while relying on known mechanisms for adaptation.

\section{Results}
In this work, we present a motor adaptation model in which the agent (e.g. a human learning a motor task) updates an existing model of the environment based on error signals produced by the need of adaptation. Several models exist for such error-based motor adaptation, but we expand on existing models by adding an explicit component of context inference. As we will show, this context inference guides the selection of the adequate internal model, its updating and the sampling of actions.

In what follows, we show that the proposed model explains several experimental phenomena found in previous studies. We simulated representative experiments from these studies to illustrate how our model explains these experimental findings using the dynamics of context inference. In what follows, we will present these simulations alongside the experimental data from the representative studies and discuss in detail how context inference explains the experimental phenomena.

@@

Central to the model is the concept of context. In our model, contexts are defined in terms of the task to be performed (see \fref{fig:model}A), the variables of the environment that are relevant to perform the task, the forward models used by the decision-making agent to perform the task, and the update mechanisms necessary to adapt these forward models to the variables of the environment. Together, these elements allow the agent to make predictions on future observations when this context is active, and these predictions are used to infer the context. For example, when lifting an object of unknown weight, an agent might have learned one context for heavy objects and one for light objects. When observing an object to be lifted, the agent can use its size and texture to estimate the weight of the object, which in turn allows the agent to infer the appropriate context and, with it, decide how to lift the object. These contextual cues (size and texture) do not necessarily identify the context (e.g. it could be a movie prop that looks like a rock) and integration over time might be necessary.

@@

It is widely accepted that in order to adapt motor commands to a novel environment, we create and update internal forward models of the outcomes of control signals \citep{Wolpert_Multiple_1998}. Here, we use this principle and chose to make use of an exact Bayesian learner. As has been discussed before \cite{Oh_Minimizing_2019}, a Bayesian learner has the advantage of not only fitting experimental data on motor adaptation, but also eliminating the need for explicit multiple time scales of learning as the learner can deal with adaptation after any number of trials. As we will show below this feature will be useful to explain specific key phenomena parsimoniously. We show that this adaptation mechanism can deal with adaptation after any number of trials, ranging from a handful of trials in an experiment to the level of expertise a professional dancer has accrued through years of practicing the same movements.

@@

Following the these ideas, our model performs context inference that integrates several sources of information, including sensory cues, proprioceptive information, outcome prediction errors and reward prediction errors.

Context is a categorical variable $\zeta$, where the identification of a context is done through Bayesian inference:
\begin{equation}
q(\zeta | \vartheta_t, \vartheta_{t-1}, c_{t-1}) \propto p(\vartheta_t | \vartheta_{t-1}, c_{t-1})p(\zeta)
\label{eqn:context-inference-mini}
\end{equation}
where the prior distribution over contexts $p(\zeta)$ includes information from two sources: (1) an expectation of continuity, encoding the expectation that the context does not change from trial to trial, and (2) an overall expectation of observing any one context, if the context did change. Below, we discuss how these two sources of information effect different experimentally observed phenomena.

@@

\section{Experimental results}
In this section, we present experimentally-observed phenomena in four sections, and show that context inference, as done in our model, provides a unifying explanation for all of them. In the first section, we discuss switches between contexts, and how slow context inference affects these switches. In the second section, we focus on the effects of volatility in the environment on motor adaptation. In the third section, we discuss the effects of slow context inference on action selection. Finally, in the fourth section we discuss context inference during error-clamp trials, and its effect on behavior.

For clarity, we first introduce necessary terminology that is typically used in experimental studies. As an example, we will use a typical motor adaptation task in which participants have to make reaching movements while holding the handle of a mechanical arm that exerts a curl force (i.e. a force that pushes the participant's hand in the direction perpendicular to movement). Depending on the trial, the arm might exert a curl force in a clockwise or counter-clockwise direction, or no force at all. Let us define the baseline context O as that in which the mechanical arm exerts no force. Context A can be defined as that in which the arm exerts the clockwise curl force and context B counter-clockwise. Additionally, abusing notation, we can say that $B = (-A)$, as the forces have the same magnitude but point in opposite directions. Studies often talk of contexts such as A and A/2, which means that both have the same direction of the adaptation (e.g. clockwise), but the second one has half the magnitude. Finally, many experiments include a block of error-clamp trials at the end of the experiment, in which the mechanical arm forces the participant to make straight-line movements; we represent these with the letter E.

@@

In this section, we discuss savings, and the related phenomenon of de-adaptation, in terms of switching between forward models. This has been modeled before \citep[e.g.][]{Wolpert_Multiple_1998,Oh_Minimizing_2019,Imamizu_Explicit_2007} but we show that expanding forward model switching into context inference explains facets of savings that these previous models do not.

To show this, we examined multiple experimental studies in which savings are observed. We categorized these studies based on the amount of contextual information made available to participants. In some experiments \citep[e.g.][]{Kim_Neural_2015,Lee_Dual_2009}, the context is clearly revealed to the participant. We call these cued-context experiments. In other experiments, partial information is available to participants \citep[e.g.][]{Davidson_Scaling_2004,Zarahn_Explaining_2008} in the form of large prediction errors, partial contextual information or reward prediction errors; we refer to these as partially-cued experiments.

@@

Because context inference integrates information from many sources, many experiments in which no intentional, overt contextual cues are available indeed contain contextual information that the participant can use to infer the context. For example, in curl-force-field experiments with mechanical arms, the force feedback provided by the mechanical arm always gives participants a sense for the current context, although error-clamp trials are difficult to distinguish from adaptation trials based on force feedback alone. In fact, the mere act of holding the handle informs participants that something is different from baseline. Proprioceptive signals can also provide contextual information \cite{Dizio_Motor_1995,Shadmehr_Adaptive_1994}. The sudden appearance of motor errors can itself be a cue for contextual change, as long as the motor error is perceivably higher than the expected trial-by-trial variation in performance, and even an unusually long pause between two trials could indicate to participants that a new block with a potentially different context could have started \cite{Ethier_Spontaneous_2008}.

To show how adaptation changes depending on the amount of contextual information, we selected three representative experiments from two studies \cite{Kim_Neural_2015,Oh_Minimizing_2019} which differ in the amount of contextual information. \cite{Kim_Neural_2015} performed a cued-context visuomotor rotation experiment with three contexts with different rotation: no rotation, 40 degrees and -40 degrees. Participants performed shooting movements in blocks of trials with the same rotation. Importantly, the identity of the current context was displayed as a visual cue (colored light), making this a cued-context experiment. Consistent with our model, the authors found that switching was immediate and accurate. In \fref{fig:oh-2019}A we show the results of simulations with our model using the parameters of the task, as well as the experimental results from \cite{Kim_Neural_2015}. The correspondence between simulations and experimental results can be seen in the switches between contexts, i.e. when the solid black line (representing the real context) switches between 40, 0 and -40; in both panels (left and right), the blue line, representing the agent's adaptation, quickly follows the switches.

@@

Participants did not completely adapt to the rotation during the first block, as evidenced by the blue line not reaching as low or high as the black line during the first blocks. Switching between opposite adaptations was nevertheless immediate in simulations and experimental data. After around 500 trials, participants had adapted completely, as seen by the blue line reaching the heights of the black line, and switches continued to be immediate.

@@

\cite{Oh_Minimizing_2019} performed partially-cued experiments similar to \cite{Kim_Neural_2015}. The authors performed two experiments that differed in the size of the adaptation: in the first experiment, the angle of rotation was 20 degrees, while in the second experiment it was only 10. The results of their experiments can be seen in \fref{fig:oh-2019}B and \ref{fig:oh-2019}C, respectively, alongside simulations with our model. Participants in the first experiment first learned the adaptation in $A_1$. From the $A_1 - O_2$ transition onward, participants showed immediate switching (with a one-trial lag) between A and O (both ways), which can be seen in terms of the blue line (adaptation) closely following the switches in the black line (real rotation). In the second experiment, switching happens more slowly, with adaptation (blue line) lagging behind the switches in real context (black line), and slowly catching up. As can be seen in the left panels in \fref{fig:oh-2019}B and \ref{fig:oh-2019}C, the same model parametrization produces fast, accurate switches when the adaptation is large, and slow, noisy switches when it is low. This difference is explained by our model in terms of the size of the adaptation: as the size of the adaptation is lower (10 degrees), it becomes more difficult to distinguish errors made by incorrectly inferring the context from the noise due to trial-to-trial variation in motor output. Because of this, the model requires more evidence (i.e. more trials) to infer a switch in contexts.

Note that the data from \cite{Oh_Minimizing_2019} and \cite{Kim_Neural_2015} include both savings (O-A transitions) and de-adaptation (A-O transitions), both of which display the same characteristics and are explained by the same mechanism in our model.

\subsection{Slower adaptation in volatile and unpredictable environments}
One central element in our model is that adaptation is modulated by the uncertainty on the current context: a prediction error observed at trial $t$ will effect motor adaptation with a magnitude proportional to the size of the error, where the proportionality constant is related to the uncertainty over the context in which the error was observed: the higher the uncertainty, the lower the size of the adaptation.

@@

\cite{Herzfeld_memory_2014} used an experimental paradigm in which the force that a mechanical arm exerts on the participant's hand can change from trial to trial between O, A and -A. Participants were divided into three groups for which the context transition from one trial to the next had differing probabilities, depending on the group: for the first group, transitions were very uncommon, for the second group they were somewhat common and for the third group they were occurred at almost every trial. Importantly, these transitions were uncued, and the participant could only infer that a transition had occurred after a motor error was observed.

@@

Our model offers a different explanation of the observed results, based on context inference: the asymmetry is due to the existence of the baseline context. As discussed above, action selection in our model is done via a weighted average (see \eref{eq:action-selection}). In experiments without cued contexts, the baseline context O and its associated forward model $m_O$, have a non-zero probability $p(\zeta | s_t ...)$. When a new block of trials starts (e.g. in the transition from 3A to A), a switch is inferred by the model (given feedback after the first trial) and $m_O$ becomes more likely (given that $m_{3A}$ has been ruled out. Therefore, in these first trials, action selection has a component guided by the baseline model, in which no extra compensatory force is applied, effectively ``pulling'' adaptation towards zero (no compensatory force). In the first group, this initial pull towards zero accelerates the transition towards A because $3A > A > 0$, but on the second group, it slows down the switch because $A > 0 > -A$.

We simulated data with our model fitting the experimental structure in \cite{Davidson_Scaling_2004}. We show the results in \fref{fig:davidson-2004}, alongside the experimental results from \cite{Davidson_Scaling_2004}. It can be seen that the agent exposed to the -8 context shifts back to 4 more slowly than the one exposed to +8, qualitatively reproducing the data from \cite{Davidson_Scaling_2004}, shown in \fref{fig:davidson-2004}B. In our simulations, the difference between the groups, as well as the variability within each group, is smaller than in the experimental data. However, the effect can be reproduced with any number of simulated runs (each one representing one participant), pointing towards a reliable effect. While we focused here on a specific, well-known experiment, we discuss experiments similar to that by \cite{Davidson_Scaling_2004} in the Discussion section.

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{./figures/figure_5.png}
\caption{Adaptation during error clamp trials. (A) Simulated adaptation during the error-clamp trials for the three groups of participants in \cite{Vaswani_Decay_2013}, using the same colors. The groups differ in the sequence of adaptations: 1.1 performed an A-E sequence; 1.2, O-A-E; 1.3, (-A/2)-A-E; and 1.4, (-A)-E. Following \cite{Vaswani_Decay_2013}, group 1.4 is not shown in A and B, as their behavior is identical to group 1.1. The solid line is the average across 10 runs (i.e. a group of 10 simulated participants) and the shaded area represents the standard deviation. The vertical dashed line is the start of the error-clamp trials. (B) Corresponding experimental data adapted from figure 2C by \cite{Vaswani_Decay_2013}. (C) Simulations: Inference over the current context, where contexts are color coded: black for baseline, orange for the counter-clockwise force, purple for the clockwise force and brown for clockwise force with half strength. The lines represent the posterior probability of each context in every trial, while the background color represents the true context. An olive-colored background represents error-clamp trials. As in (A), solid lines represent the average across all runs and shaded areas represent the standard deviation. (D) Simulations: Visualization of the lag before a change in context is detected by the agent during the E trials. Each line represents one run (10 runs per group).}
\label{fig:vaswani-2013}
\end{figure}

@@

The same rationale explains the results from \cite{Davidson_Scaling_2004}: motor learning, which in our model is modulated by context inference, is minimal for errors close to 2 and -2 (see their figure 2E). This is because an error of 2 or -2 signals that the participant incorrectly identified the context (as adaptation has a magnitude of 1).

>>>>

\setlength{\parskip}{1em}
\def \fref #1{Figure \ref{#1}}     % Reference figures
\def \tref #1{Table \ref{#1}}      % Reference tables
\def \eref #1{Equation \ref{#1}}   % Reference equations
\def \sref #1{Section '\nameref{#1}'}    % Reference sections

@@

\abstract{Abstract}
Humans have been shown to adapt their movements when a sudden change to the dynamics of the environment is introduced, a phenomenon called motor adaptation. If the change is reverted, the adaptation is also quickly reverted. Human are also able to adapt to multiple changes in dynamics presented separately, and to be able to switch between adapted movements on the fly. Such switching relies on contextual information which is often noisy or misleading, which affects the switch between adaptations. In this work, we use a computational mechanistic model to explain the behavioral phenomena effected by imperfect contextual information. Specifically, we present a hierarchical computational model for motor adaptation based on exact Bayesian inference. This model explicitly takes into account contextual information and how the dynamics of context inference affect learning and action selection. We show how the proposed model provides a unifying explanation for four different experimentally established phenomena: (i) effects of sensory cues and proprioceptive information on switching between tasks, (ii) the effects of environmental volatility and predictability on learning, (iii) the effects of previously-learned adaptations on switching between tasks, in addition to  (iv) the well studied savings, de-adaptation and spontaneous recovery.

\section{Introduction}
Humans are able to adapt their movements to changing dynamics of the environment and of the body itself. We adapt to slow changes in our body dynamics, e.g. because of muscle exhaustion throughout the day, but we can also adapt to sudden changes in the environment. For example, when picking up a heavy object, we must adapt our motor commands to be able to move the extra weight safely. Even before picking it up, the size and material of an object provide us with contextual information which we can use to prepare the correct posture for moving the object. However, contextual information cannot always be relied on, e.g. a large parcel with unknown content.

It has been shown that humans can learn multiple tasks during the course of a single experiment, as well as learn to adapt to often-opposing modifications to the dynamics of movements \citep{Gandolfo_Motor_1996,Shadmehr_Functional_1997}, although interference between tasks can slow down learning \citep{Brashers-Krug_Consolidation_1996,Sing_Reduction_2010}. Additionally, humans have been shown to dynamically switch between different learned adaptation \citep{Davidson_Scaling_2004,Ethier_Spontaneous_2008,Lee_Dual_2009}. The mechanisms behind these switches, the specifics of how these switches manifest in behavior, and the ways in which switching can be impeded or improved, are an ongoing, active topic of research.

@@

By introducing blocks of trials in which a manipulation is present (e.g. a mechanical arm exerts a force), experimenters are able to observe the dynamics of adaptation through the lens of motor error. Across many different motor adaptation experiments, now well-established phenomena have been observed, for example: (i) savings (the ability to recall previously-learned skills), (ii) quick de-adaptation (the ability to return to unmodified dynamics), (iii) anterograde interference (the interference in motor learning between opposing manipulations in dynamics) and (iv) spontaneous recovery (during trials where errors are forced to be zero, the display of behavior consistent with the previously-learned dynamics).

To explain these phenomena, a number of computational models have been introduced. The least complex of these models are linear learners \citep{Smith_Interacting_2006,Forano_Timescales_2020,Scheidt_Learning_2001}, in which a single one-dimensional variable is used to describe the adaptation of motor commands given the observed motor errors. By having multiple time scales, each with a learning rate and a forgetting rate, these models have been fitted to experimental data, partially explaining these distinct experimental phenomena.

Bayesian accounts of motor adaptation have also been presented, providing an alternative explanation for savings and quick de-adaptation in the form of switching between internal models of adaptation via predictions generated by these internal models \citep{Kording_Bayesian_2004,Oh_Minimizing_2019}. These models replace the idea of explicit multiple time scales with an estimation of their own uncertainty: as the the same task is practiced, the uncertainty on an estimate of the underlying modification is reduced, which lowers variability in responses and increases the stability of the estimate.

@@

In this work, we show that these unexplained phenomena, alongside savings, de-adaptation, anterograde interference and spontaneous recovery, can be explained in terms of the concept of context, an idea that has recently been gaining traction in different fields \cite[e.g.][]{Sanders_Hippocampal_2020,Hunter_Contextsensitive_2021}. Intuitively, in mechanical arm experiments, one can think of one context in which the arm exerts no force and one in which it does. By inferring the current context, a participant can know how to counteract the mechanical arm and succeed in a reaching movement. In some experiments, the context is clearly inferable via sensory cues \cite[e.g.][]{Gandolfo_Motor_1996,Shadmehr_Adaptive_1994,Ethier_Spontaneous_2008}. In many other experiments, the context is not directly observable, and context inference is probabilistic and takes the form of an evidence-accumulating process. In this work, we propose that the behavioural consequences of this slow process of evidence accumulation can be observed in behavior and can be used to explain the phenomena outlined above.

To this end, we present a computational model for motor adaptation inspired by models such as MOSAIC \citep{Wolpert_Multiple_1998} and Bayesian switching models \citep{Kording_Bayesian_2004,Oh_Minimizing_2019}. We build on these models by expanding the concept of context and introducing an explicit mechanism for inferring the context based on Bayesian inference. In our model, the context encompasses all the information relevant to making decisions, both environmental and internal to the agent making these decisions. To infer the current context, our model integrates several sources of information, including the predictions made by its internal forward models, an estimate of the reliability of these predictions, direct sensory information (such as visual contextual cues), proprioceptive information, and feedback provided by the environment (e.g. whether the task was successfully completed). Having inferred the context, the model can choose the best-fitting internal model with which to make decisions.

We show that the dynamics of context inference provide a principled mechanistic explanation for the experimental phenomena outlined above, providing a single account for motor adaptation under changing contexts, while relying on known mechanisms for adaptation.

\section{Results}
In this work, we present a motor adaptation model in which the agent (e.g. a human learning a motor task) updates an existing model of the environment based on error signals produced by the need of adaptation. Several models exist for such error-based motor adaptation \cite[e.g.][]{Wolpert_Multiple_1998,Kording_Bayesian_2004,Oh_Minimizing_2019}. We expand on these existing models by adding an explicit component of context inference. As we will show, this context inference guides the selection of the adequate internal model, its updating and the sampling of actions.

In what follows, we show that the proposed model explains several experimental phenomena found in previous studies. We simulated representative experiments from these studies to illustrate how our model explains these experimental findings using the dynamics of context inference. We will present these simulations alongside the experimental results from the representative studies and discuss in detail how context inference explains the experimental phenomena.

@@

Central to the model is the concept of context. In our model, contexts are defined in terms of the task to be performed (see \fref{fig:model}A), the variables of the environment that are relevant to perform the task, the forward models used by the decision-making agent to perform the task, and the update mechanisms necessary to adapt these forward models to the variables of the environment. Together, these elements allow the agent to make predictions on future observations when this context is active, and these predictions are used to infer the context. For example, when lifting an object of unknown weight, an agent might have learned one context for heavy objects and one for light objects. When observing an object to be lifted, the agent can use its size and texture to estimate the weight of the object, which in turn allows the agent to infer the appropriate context and, with it, decide how to lift the object. These contextual cues (size and texture) do not necessarily identify the context with certainty (e.g. it could be a movie prop that looks like a rock) and integration over time might be necessary.

@@

It is widely accepted that in order to adapt motor commands to a novel environment, we create and update internal forward models of the outcomes of control signals \citep{Wolpert_Multiple_1998}. Here, we use this principle and chose to make use of an exact Bayesian learner. As has been discussed before \cite{Oh_Minimizing_2019}, a Bayesian learner has the advantage of not only fitting experimental data on motor adaptation, but also eliminating the need for explicit multiple time scales of learning as the learner can deal with adaptation after any number of trials. As we will show below this feature will be useful to explain specific key phenomena parsimoniously.

@@

Following these ideas, our model performs context inference that integrates several sources of information, including sensory cues, proprioceptive information, outcome prediction errors and reward prediction errors.

In the model, context is a categorical variable $\zeta$, where the identification of a context is done through Bayesian inference:
\begin{equation}
q(\zeta | \vartheta_t, \vartheta_{t-1}, c_{t-1}) \propto p(\vartheta_t | \vartheta_{t-1}, c_{t-1})p(\zeta)
\label{eqn:context-inference-mini}
\end{equation}
where the prior distribution over contexts $p(\zeta)$ includes information from two sources: (1) an expectation of continuity, encoding the expectation that the context does not change from trial to trial, and (2) an overall expectation of observing any one context, if the context did change. Below, we discuss how these two sources of information effect different experimentally observed phenomena.

@@

\section{Experimental results}
In this section, we present experimentally-observed phenomena in four sections, and show that context inference, as done in our model, provides a unifying explanation for all of them. In the first section, we discuss switches between contexts, and how slow context inference affects these switches. In the second section, we focus on the effects of volatility in the environment on motor adaptation. In the third section, we discuss the effects of slow context inference on action selection. Finally, in the fourth section we discuss context inference during so-called error-clamp trials, and its effect on behavior.

For clarity, we first introduce necessary terminology that is typically used in experimental studies and will be useful for linking experimental to computational observations. As an example, we will use a typical motor adaptation task in which participants have to make reaching movements while holding the handle of a mechanical arm that exerts a curl force (i.e. a force that pushes the participant's hand in the direction perpendicular to movement). Depending on the trial, the mechanical arm might exert a curl force in a clockwise or counter-clockwise direction, or no force at all. Let us define the baseline context O as that in which the mechanical arm exerts no force. Context A can be defined as that in which the arm exerts the clockwise curl force and context B counter-clockwise. Additionally, abusing notation, the usual statement is that $B = (-A)$, as the forces have the same magnitude but point in opposite directions. Studies often talk of contexts such as A and A/2, which means that both have the same direction of the adaptation (e.g. clockwise), but the second one has half the magnitude. Finally, many experiments include a block of error-clamp trials at the end of the experiment, in which the mechanical arm forces the participant to make straight-line movements; we represent these with the letter E.

@@

In this section, we discuss savings, and the related phenomenon of de-adaptation, in terms of switching between forward models. This has been modeled before \citep[e.g.][]{Wolpert_Multiple_1998,Oh_Minimizing_2019,Imamizu_Explicit_2007} where we show in addition that expanding forward model switching into context inference explains facets of savings that these previous models do not.

To show this, we examined multiple experimental studies in which savings are observed. We categorized these studies based on the amount of contextual information made available to participants into cued-context and partially-cued experiments: In some experiments \citep[e.g.][]{Kim_Neural_2015,Lee_Dual_2009}, the context is clearly revealed to the participant. We call these cued-context experiments. In other experiments, partial information is available to participants \citep[e.g.][]{Davidson_Scaling_2004,Zarahn_Explaining_2008} in the form of large prediction errors, partial contextual information or reward prediction errors; we refer to these as partially-cued experiments.

@@

Because context inference integrates information from different sources, many experiments in which no intentional, overt contextual cues are available indeed contain contextual information that the participant can use to infer the context. For example, in curl-force-field experiments with mechanical arms, the force feedback provided by the mechanical arm always gives participants a sense for the current context, although error-clamp trials are difficult to distinguish from adaptation trials based on force feedback alone. In fact, the mere act of holding the handle informs participants that something is different from baseline. Proprioceptive signals can also provide contextual information \cite{Dizio_Motor_1995,Shadmehr_Adaptive_1994}. The sudden appearance of motor errors can itself be a cue for contextual change, as long as the motor error is perceivably higher than the expected trial-by-trial variation in performance, and even an unusually long pause between two trials could indicate to participants that a new block with a potentially different context could have started \cite{Ethier_Spontaneous_2008}.

To show how adaptation changes depending on the amount of contextual information, we selected three representative experiments from two studies \cite{Kim_Neural_2015,Oh_Minimizing_2019} which differ in the amount of contextual information. \cite{Kim_Neural_2015} performed a cued-context visuomotor rotation experiment with three contexts with different rotation: no rotation, 40 degrees and -40 degrees. Participants performed shooting movements in blocks of trials with the same rotation. Importantly, the identity of the current context was displayed as a visual cue (colored light), making this a cued-context experiment. Consistent with our model, the authors found that switching was immediate and accurate. In \fref{fig:oh-2019}A we show the results of simulations with our model using the parameters of the task, as well as the experimental results from \cite{Kim_Neural_2015}. The correspondence between simulations and experimental results can be seen in the switches between contexts, i.e. when the solid black line (representing the true context) switches between 40, 0 and -40; in both panels (left and right), the blue line, representing the agent's adaptation, quickly follows the switches.

@@

Critically, in \fref{fig:oh-2019}A it can be seen until trial 250 and especially for the first four context switches,  that participants did not completely adapt to the rotation, as evidenced by the blue line not reaching as low or high as the black line during the first blocks. Switching between opposite adaptations was nevertheless immediate in simulations and experimental data. After around 500 trials, participants had adapted completely, as seen by the blue line reaching the heights of the black line, and switches continued to be immediate.

@@

\cite{Oh_Minimizing_2019} performed partially-cued experiments similar to \cite{Kim_Neural_2015}. The authors performed two experiments that differed in the size of the adaptation: in the first experiment, the angle of rotation was 20 degrees, while in the second experiment it was only 10. The results of their experiments can be seen in \fref{fig:oh-2019}B and \ref{fig:oh-2019}C, respectively, alongside simulations with our model. Participants in the first experiment (\fref{fig:oh-2019}B) first learned the adaptation in $A_1$. From the $A_1 - O_2$ transition onward, participants showed immediate switching (with a one-trial lag) between A and O (both ways), which can be seen in terms of the blue line (adaptation) closely following the switches in the black line (true rotation). In the second experiment (\fref{fig:oh-2019}C), switching happens more slowly, with adaptation (blue line) lagging behind the switches in the real context (black line), and slowly catching up. As can be seen in the left panels in \fref{fig:oh-2019}B and \ref{fig:oh-2019}C, the same model parametrization produces fast, accurate switches when the adaptation is large (B), and slow, noisy switches when it is low (C). This difference is explained by our model in terms of the size of the adaptation: as the size of the adaptation is lower (10 degrees), it becomes more difficult to distinguish errors made by incorrectly inferring the context from the noise due to trial-to-trial variation in motor output. Because of this, the model requires more evidence (i.e. more trials) to infer a switch in contexts.

Note that the results from our simulations, from \cite{Oh_Minimizing_2019} and from \cite{Kim_Neural_2015} include both savings (O-A transitions) and de-adaptation (A-O transitions), both of which display the same characteristics and are explained by the same mechanism in our model.

\subsection{Slower adaptation in volatile and unpredictable environments}
One central element in the proposed model is that adaptation is modulated by the uncertainty on the current context: a prediction error observed at trial $t$ will effect motor adaptation with a magnitude proportional to the size of the error, where the proportionality constant is related to the uncertainty over the context in which the error was observed: the higher the uncertainty, the lower the size of the adaptation.

@@

\cite{Herzfeld_memory_2014} used an experimental paradigm in which the force that a mechanical arm exerts on the participant's hand can change from trial to trial between O, A and -A. Participants were divided into three groups for which the context transition from one trial to the next had different probabilities, depending on the group: for the first group, transitions were very uncommon, for the second group they were somewhat common and for the third group they occurred at almost every trial. Importantly, these transitions were uncued, and the participant could only infer that a transition had occurred after a motor error was observed.

@@

Our model offers a different explanation of the observed results, based on context inference: the asymmetry is due to the existence of the baseline context. As discussed above, action selection in our model is done via a weighted average (see \eref{eq:action-selection}). In experiments without cued contexts, the baseline context O and its associated forward model $m_O$, have a non-zero probability $p(\zeta | s_t ...)$. When a new block of trials starts (e.g. in the transition from 3A to A), a switch is inferred by the model (given feedback after the first trial) and $m_O$ becomes more likely (given that $m_{3A}$ has been ruled out). Therefore, in these first trials, action selection has a component guided by the baseline model, in which no extra compensatory force is applied, effectively ``pulling'' adaptation towards zero (no compensatory force). In the first group, this initial pull towards zero accelerates the transition towards A because $3A > A > 0$, but in the second group, it slows down the switch because $A > 0 > -A$.

We simulated data with our model fitting the experimental structure in \cite{Davidson_Scaling_2004}. We show the results in \fref{fig:davidson-2004}, alongside the experimental results from \cite{Davidson_Scaling_2004}. It can be seen that the agent exposed to the -8 context shifts back to 4 more slowly than the one exposed to +8, qualitatively reproducing the data from \cite{Davidson_Scaling_2004}, shown in \fref{fig:davidson-2004}B. Note that in our simulations, the difference between the groups, as well as the variability within each group, is smaller than in the experimental data. However, the effect can be reproduced with any number of simulated runs (each one representing one participant), pointing towards a reliable effect. While we focused here on a specific, well-known experiment, we discuss experiments similar to that by \cite{Davidson_Scaling_2004} in the Discussion.

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{./figures/figure_5.png}
\caption{Adaptation during error clamp trials. (A) Simulated adaptation during the error-clamp trials for the three groups of participants in \cite{Vaswani_Decay_2013}, using the same colors. The groups differ in the sequence of adaptations: 1.1 performed an A-E sequence; 1.2, O-A-E; 1.3, (-A/2)-A-E; and 1.4, (-A)-E. Following \cite{Vaswani_Decay_2013}, group 1.4 is not shown in A and B, as their behavior is identical to group 1.1. The solid line is the average across 10 runs (i.e. a group of 10 simulated participants) and the shaded area represents the standard deviation. The vertical dashed line is the start of the error-clamp trials. (B) Corresponding experimental data adapted from figure 2C by \cite{Vaswani_Decay_2013}. (C) Simulations: Inference over the current context, where contexts are color coded: black for baseline, orange for the counter-clockwise force, purple for the clockwise force and brown for counter-clockwise force with half strength. The lines represent the posterior probability of each context in every trial, while the background color represents the true context. An olive-colored background represents error-clamp trials. As in (A), solid lines represent the average across all runs and shaded areas represent the standard deviation. (D) Simulations: Visualization of the lag before a change in context is detected by the agent during the E trials. Each line represents one run (10 runs per group).}
\label{fig:vaswani-2013}
\end{figure}

@@

The same rationale explains the results by \cite{Davidson_Scaling_2004}: motor learning, which in our model is modulated by context inference, is minimal for errors close to 2 and -2 (see their figure 2E). This is because an error of 2 or -2 signals that the participant incorrectly identified the context (as adaptation has a magnitude of 1).
